---
title: "Computing Assignment – Bayesian phylogenetics and epidemiology"
output:
  html_document: default
  pdf_document: default
---

```{r setup, message=FALSE, warning=FALSE, results='hide'}
# List of required packages
required_packages <- c(
  "rmarkdown", "tinytex", "tidyverse", "ggtree", 
  "ggridges", "treeio", "coda", "cols4all")

# Install missing packages by comparing against the currently installed packages
missing_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if(length(missing_packages)) {
  install.packages(missing_packages)
}
#loads all the packages by applying the library function to each
lapply(required_packages, library, character.only = TRUE)
```

Candidate Number: 1073601

**Question 1**

From the trees produced in the R Markdown file propose how many independent CHIKV introductions into Kassala sparked and sustained the 2018 outbreak, and their time of occurrence. How much earlier from the onset of the current outbreak is this estimated to be, and what does this imply? Use tree visualisations and the distribution of relevant parameter estimates to explain your answer.

```{r}
# Load the MCC tree
beast_file <- ("MCC_trees/MCC_CHIKV_IndianOcean.tree")
beast_tree <- treeio::read.beast(beast_file)

# Plots a tree showing the latest possible Kassala MRCA 
ggtree(beast_tree, aes(color = outbreak), mrsd="2018-10-16") + theme_tree2() +
  scale_x_continuous(breaks = c(2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016, 2018),
                     minor_breaks = seq(2004, 2019, 1)) +
  geom_point2(aes(subset = (node == 255)), shape = 21, size = 3, fill = 'darkblue') +
  scale_color_discrete_c4a_cat("friendly7") +
  theme(panel.grid.major = element_line(color = "black", size = .1),
        panel.grid.minor = element_line(color = "grey", size = .1),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())
#Scale bars are added, and the outbreaks are represented using colour-blind friendly visuals
```

As you can see from the above graph, all the Kassala nodes are monophyletic and branch from the same individual MRCA which suggests that only one independent CHIKV introduction caused the 2018 outbreak. The estimated 
